Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-08-25 manuscript/manuscript.tex
Modifications on manuscript

In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.
We have performed similar tests, but now with variable density spherical shell.
From the infinite density functions that can be proposed, we have chosen three characteristic ones: a linear, an exponential and a discontinuous function.
The first one constitutes the simplest variation, while the second one represents the fastest varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

@@

In order to test the new variable density tesseroid code and estimate the error of the approximation we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
This error estimations are also useful in the determination of the minimum value for the distance-size ratio $D$ such the difference between the numerical model and the analytical solutions are below an acceptable threshold.

@@

We computed the three gravity fields for different values of $D$, ranging from 0.5 to 10 with a step of 0.5, on every point of these four grids.
Then we calculated the maximum difference between these results and the value of the analytical solution at the same height.

@@

The tests were only performed on these magnitudes because the other components of the gravity gradient and Marussi tensor are equal to zero due to the rotational symmetries of the spherical shell.

@@

As a simple case, we proposed a linear density function in such a way that the spherical shell has a density of 2670~kg/m$^3$ on the surface of the Earth, and 3300~kg/m$^3$ at 35 km of depth:

\begin{equation}
    \rho(r') = ar' + c,
    %~ -\frac{3300 - 2670}{35000} r' + 
    %~ \left( \frac{3300 - 2670}{35000} R + 2670 \right)
    \label{eq:density-linear}
\end{equation}
\noindent with 
\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{35000\text{m}},
\end{equation}
\begin{equation}
    c = \frac{3300\text{kg/m$^3$} - 
        2670\text{kg/m$^3$}}{35000\text{m}} R + 
        2670\text{kg/m$^3$},
\end{equation}

@@

The tests were performed on a thin and a thick spherical shell of 1 km and 35 km of thickness, respectively.
In Figures \ref{fig:D-linear-thin} and \ref{fig:D-linear-thick} we can see the results for the potential, the vertical component of the gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor.
Because the other two diagonal components of the tensor show similar results as the ones obtained for the $g_{zz}$, we exclude them from the figure, although they can be found in the repository. \todo{chequear}

@@

\noindent where $A$ = 1kg/m$^3$ and $b$ = 1000km.
We have also performed this test with a thin and a thick shell of 1 km and 35 km, respectively, and the results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

@@

Because the constant $b$ controls the variation of the density function we wanted to test how does the accuracy of the method behaves at different values of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model and the analytical solution on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}).
For each value of $b$ we calculated these difference for the same set of values of $D$ we've been using on the previous tests, obtaining one curve for each kind of variation.

This tests were performed using both the thin and the thick spherical shell, and the results can be seen in Figure \ref{fig:D-exp-power}.
Taking into account that lower values of $b$ implies a more varying function, we can see that for extremely varying functions the numerical model does not perform well: for $b$=100m the difference is almost 100\% in the thin shell case, and the same happens with $b$=10km in the thick shell case.

@@

Secondly, lets assume the same density function from equation \ref{eq:density-discontinuous}, but now with an asymmetrical distribution: $R_c = (R_1 + R_2)/3$, and perform the same test.
We can see its results in Figure \ref{fig:D-discont-asym}: up to $D=10$ the difference between the numerical model and the analytical solution is greater than 10\%, what
leads us to conclude that the code does not reassembles the gravity fields for this case with good accuracy.

@@

We have recorded the computation times for the gravity potential, the gradient component $g_z$ and the Marussi tensor component $g_{zz}$ generated by a single tesseroid model with homogeneous and variable density at different heights (see Figure \ref{fig:speed-comparison}).
Firstly, we can see that both codes take more time to compute the gravity field for lower heights due to the increased divisions of the tesseroid in such cases.
Secondly, we can see that the homogeneous density code is up to twice faster than the new variable density algorithm for lower heights, however the computation times gets more similar values at greater heights. In conclusion, the new variable dnesity code is fast enough to perform the calculations of a real model in a few minutes. \todo{real, esta bien la expresion?}

>>>>

In order to overcome this, \citet{Uieda2016} compared the numerical computation with the analytical solution for a spherical shell with homogeneous density.
We have performed similar tests, but now with variable density spherical shell.
From the infinite density functions that can be proposed, we have chosen three characteristic ones: a linear, an exponential and a discontinuous function.

@@

In order to test the new variable density tesseroid code and estimate the error of the approximation, we have compared the numerical model with the analytical solutions for a spherical shell with variable density for three special cases: a linear, an exponential and a discontinuous density function.
The first one constitutes the simplest variation, while the second one represents the fastest varying function. The later one tries to emulate a contact between two different media (e.g. crust and mantle interface).

This error estimations are also useful in the determination of the minimum value for the distance-size ratio $D$ such the difference between the numerical model and the analytical solutions are below an acceptable threshold.

@@

We computed the three gravity fields  on every point of these four grids for different values of $D$, ranging from 0.5 to 10 with a step of 0.5.
Then we calculated the maximum difference between these results and the value of the analytical solution at the same height.

@@

@@

In first place, we proposed a spherical shell with linear density in such way that it has a density of 2670~kg/m$^3$ on the surface of the Earth, and 3300~kg/m$^3$ at 35 km of depth:

\begin{equation}
    \rho(r') = ar' + c,
    \label{eq:density-linear}
\end{equation}
\noindent with 
\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{35000\text{m}},
\end{equation}
\begin{equation}
    c = \frac{3300\text{kg/m$^3$} - 
        2670\text{kg/m$^3$}}{35000\text{m}} R + 
        2670\text{kg/m$^3$},
\end{equation}

@@

Then we computed the difference between the numerical model with its corresponding analytical solution (equation \ref{eq:density-linear}) for a thick and a thin spherical shell of 1km and 35km of thickness, respectively.

In Figures \ref{fig:D-linear-thin} and \ref{fig:D-linear-thick} we can see the results for the potential, the vertical component of the gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor.
Because the other two diagonal components of the tensor show similar results as the ones obtained for the $g_{zz}$, we exclude them from the figure, although they can be found in the repository. \todo{chequear}

@@

\noindent where $A$ = 1kg/m$^3$ and $b$ = 1000km.
We have performed this test also with a thin and a thick shell of 1 km and 35 km, respectively, whose results can be seen in Figures \ref{fig:D-exp-thin} and \ref{fig:D-exp-thick}.
Once again, the results for $g_{xx}$ and $g_{yy}$ are very similar to the ones obtained for $g_{zz}$, so we exclude them, although, they can be found in the repository.

@@

Because the constant $b$ controls the variation of the density function, we wanted to test how does the accuracy of the method behaves at different values of $b$.
We computed the difference between the Marussi tensor component $g_{zz}$ generated by the numerical model on the big grid of 30$^\circ\times$30$^\circ$ (see Table \ref{tab:grids}) and the analytical solution.
For each value of $b$ we calculated these difference for the same set of values of $D$ we've been using on the previous tests, obtaining one curve for each kind of variation.

This tests were performed using both the thin and the thick spherical shell, whose results can be seen in Figure \ref{fig:D-exp-power}.
Taking into account that lower values of $b$ implies a more varying function, we can see that for extremely varying functions the numerical model does not perform well: for $b$=100m the difference is almost 100\% in the thin shell case, and the same happens with $b$=10km in the thick shell case.

@@

Secondly, lets assume the same density function from equation \ref{eq:density-discontinuous}, but now with an asymmetrical distribution: $R_c = (R_1 + R_2)/3$, and perform the same test.
We can see its results in Figure \ref{fig:D-discont-asym}: up to $D=10$ the difference between the numerical model and the analytical solution is greater than 10\%, what
leads us to conclude that the code does not recreates the gravity fields for this case with good accuracy.

@@

We have recorded the computation times for the gravity potential, the gradient component $g_z$ and the Marussi tensor component $g_{zz}$ generated by a single tesseroid model with homogeneous and variable density at different heights (see Figure \ref{fig:speed-comparison}).
Firstly, we can see that both codes take more time to compute the gravity field for lower heights due to the increased divisions of the tesseroid in such cases.
Secondly, we can see that the homogeneous density code is up to twice faster than the new variable density algorithm for lower heights. However, the computation times gets more similar values at greater heights. In conclusion, the new variable density code is fast enough to perform the calculations of a real model in a few minutes. \todo{real, esta bien la expresion?}
